MATH/COMP 562 ASSIGNMENT 3

### 2023-06-24 assignment-3/article.tex
Add nested enumeration
New page split
Solve 3.6
Write up random variable question

\documentclass[
10pt, % Main document font size
a4paper, % Paper type, use 'letterpaper' for US Letter paper
oneside, % One page layout (no page indentation)
%twoside, % Two page layout (page indentation for binding and different headers)
headinclude,footinclude, % Extra spacing for the header and footer
BCOR5mm, % Binding correction
]{scrartcl}

\usepackage{mdframed}
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{argmax} % thin space, limits underneath in displays
\DeclareMathOperator*{\argmin}{argmin} % thin space, limits underneath in displays
\DeclareMathOperator*{\sig}{sig}

@@

\begin{solution}
  \begin{enumerate}[label=(\alph*)]
      \item We begin by calculating the left hand side of the strong Jensen's inequality:
        \begin{align*}
        &\hphantom{=} ||\alpha w + (1-\alpha)u)||^2 + \alpha(1-\alpha)||w-u||^2 \\
        &= (\alpha w + (1-\alpha)u)^T(\alpha w + (1-\alpha)u) + \alpha(1-\alpha)(w-u)^T(w-u) \\
        &= \alpha^2||w||^2 + 2\alpha(1-\alpha)(u\cdot w) + (1-\alpha)^2||u||^2 + \alpha(1-\alpha)(||w||^2 + 2(u\cdot w) + ||u||^2) \\
        &= (\alpha^2+\alpha(1-\alpha))||w||^2 + 4\alpha(1-\alpha)(u\cdot w) + ((1-\alpha)^2 + \alpha(1-\alpha))||u||^2 \\
        &\leq \alpha||w||^2 + 4\alpha(1-\alpha)||u||\cdot ||w|| + (1-\alpha)||u||^2 \\
        &\leq \alpha||w||^2 + (1-\alpha)||u||^2
        \end{align*}
        as required.
      \item I claim that $f(w)=|w|$ is convex, but not $\lambda$-strongly convex for any $\lambda$. By the triangle inequality, the following inequality is tight:
        \begin{align*}
          0 &\leq \alpha|w| + (1-\alpha)|u| -|\alpha w + (1-\alpha)u)|
        \end{align*}
        Strong convexity requires we can bound this term below by something quadratic in $||w-u||$ but this is not possible, so $|w|$ is not strongly convex.
      \item We calculate Jensen's inequality for $g$: \begin{align*}
        &\hphantom{=} g(\alpha w + (1-\alpha)u, \lambda) + \frac{\lambda}{2}\alpha(1-\alpha)||w-u||^2 \\
        &= f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}||\alpha w + (1-\alpha) u||^2 + \frac{\lambda}{2}\alpha(1-\alpha)||w-u||^2 \\
        &\leq f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}(\alpha||w||^2 + (1-\alpha)||u||^2 - \alpha(1-\alpha)||w-u||^2) + \frac{\lambda}{2}\alpha(1-\alpha)||w-u||^2 \\
        &\leq f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}(\alpha||w||^2 + (1-\alpha)||u||^2) \\
        &\leq \alpha (f(w) + \frac{\lambda}{2}||w||^2) + (1-\alpha)(f(u) + \frac{\lambda}{2}||u||^2)
      \end{align*}
        as required.
      \item Suppose $f(w) = g(w) - \frac{\lambda}{2}||w||^2$ is convex. We will show that $g$ is $\lambda$-strongly convex:
        \begin{align*}
          &\hphantom{=} g(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}\alpha(1-\alpha)||w-u||^2 \\
          &= f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}||\alpha w + (1-\alpha) u||^2 + \frac{\lambda}{2}\alpha(1-\alpha)||w-u||^2
        \end{align*}
  \end{enumerate}
\end{solution}

\begin{enumerate}
\setcounter{enumi}{5}
    \item Exercise 3.6 (Random Variables and Strong Convexity). Let $X \in \mathbb{R}$ be a discrete vector-valued random variable, with distribution
    \[
    \rho(X) =
    \begin{cases}
      w, & \text{with probability }\alpha \\
      u, & \text{with probability }(1 - \alpha)
    \end{cases}
    \]
    \begin{enumerate}
        \item Find $E[X]^2$, $E[X^2]$, $Var(X)$. Show that the expression $E[X^2] - (E[X])^2 = Var(X)$ corresponds to
        \[
        \alpha f(w) + (1 - \alpha) f(u) - f(\alpha w + (1 - \alpha) u) = \alpha (1 - \alpha) \|w - u\|^2
        \]
        with $f(w) = w^2$.

        \item Suppose $f$ is $\lambda$ strongly convex. Let $X$ be as above. Show that
        \[
        E[f(X)] - f(E[X]) \geq \frac{\lambda}{2} Var(X)
        \]

>>>>

\documentclass[
10pt, % Main document font size
a4paper, % Paper type, use 'letterpaper' for US Letter paper
oneside, % One page layout (no page indentation)
%twoside, % Two page layout (page indentation for binding and different headers)
headinclude,footinclude, % Extra spacing for the header and footer
BCOR5mm, % Binding correction
]{scrartcl}
\renewcommand{\labelenumii}{\theenumii}
\renewcommand{\theenumii}{\theenumi.\arabic{enumii}.}
\usepackage{mdframed}
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{argmax} % thin space, limits underneath in displays
\DeclareMathOperator*{\argmin}{argmin} % thin space, limits underneath in displays
\DeclareMathOperator*{\sig}{sig}

@@

\begin{solution}
  \begin{enumerate}[label=(\alph*)]
      \item We begin by calculating the left hand side of the strong Jensen's inequality:
        \begin{align*}
        &\hphantom{=} ||\alpha w + (1-\alpha)u)||^2 + \alpha(1-\alpha)||w-u||^2 \\
        &= (\alpha w + (1-\alpha)u)^T(\alpha w + (1-\alpha)u) + \alpha(1-\alpha)(w-u)^T(w-u) \\
        &= \alpha^2||w||^2 + 2\alpha(1-\alpha)(u\cdot w) + (1-\alpha)^2||u||^2 + \alpha(1-\alpha)(||w||^2 + 2(u\cdot w) + ||u||^2) \\
        &= (\alpha^2+\alpha(1-\alpha))||w||^2 + 4\alpha(1-\alpha)(u\cdot w) + ((1-\alpha)^2 + \alpha(1-\alpha))||u||^2 \\
        &\leq \alpha||w||^2 + 4\alpha(1-\alpha)||u||\cdot ||w|| + (1-\alpha)||u||^2 \\
        &\leq \alpha||w||^2 + (1-\alpha)||u||^2
        \end{align*}
        as required.
      \item I claim that $f(w)=|w|$ is convex, but not $\lambda$-strongly convex for any $\lambda$. By the triangle inequality, the following inequality is tight:
        \begin{align*}
          0 &\leq \alpha|w| + (1-\alpha)|u| -|\alpha w + (1-\alpha)u)|
        \end{align*}
        Strong convexity requires we can bound this term below by something quadratic in $||w-u||$ but this is not possible, so $|w|$ is not strongly convex.
      \item We calculate Jensen's inequality for $g$: \begin{align*}
        &\hphantom{=} g(\alpha w + (1-\alpha)u, \lambda) + \frac{\lambda}{2}\alpha(1-\alpha)||w-u||^2 \\
        &= f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}||\alpha w + (1-\alpha) u||^2 + \frac{\lambda}{2}\alpha(1-\alpha)||w-u||^2 \\
        &\leq f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}(\alpha||w||^2 + (1-\alpha)||u||^2 - \alpha(1-\alpha)||w-u||^2) + \frac{\lambda}{2}\alpha(1-\alpha)||w-u||^2 \\
        &\leq f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}(\alpha||w||^2 + (1-\alpha)||u||^2) \\
        &\leq \alpha (f(w) + \frac{\lambda}{2}||w||^2) + (1-\alpha)(f(u) + \frac{\lambda}{2}||u||^2)
      \end{align*}
        as required.
        \newpage
      \item Suppose $f(w) = g(w) - \frac{\lambda}{2}||w||^2$ is convex. We will show that $g$ is $\lambda$-strongly convex:
        \begin{align*}
          &\hphantom{=} g(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}\alpha(1-\alpha)||w-u||^2 \\
          &= f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}||\alpha w + (1-\alpha) u||^2 + \frac{\lambda}{2}\alpha(1-\alpha)||w-u||^2 \\
          &= f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}(\alpha w + (1-\alpha) u)^T(\alpha w + (1-\alpha) u) + \\
          &\hphantom{=} \frac{\lambda}{2}\alpha(1-\alpha)(w-u)^T(w-u) \\
          &= f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}\left(\alpha^2||w||^2+2\alpha(1-\alpha)(u\cdot w) + (1-\alpha)^2||u||^2\right) + \\
          &\hphantom{=} \frac{\lambda}{2}\alpha(1-\alpha)\left(||w||^2 - 2(u\cdot w) + ||u||^2\right) \\
          &= f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}\left(\alpha^2||w||^2 + (1-\alpha)^2||u||^2\right) + \frac{\lambda}{2}\alpha(1-\alpha)\left(||w||^2 + ||u||^2\right) \\
          &= f(\alpha w + (1-\alpha)u) + \frac{\lambda}{2}||w||^2\left(\alpha^2 + \alpha(1-\alpha)\right) + \frac{\lambda}{2}||u||^2\left((1-\alpha)^2 + \alpha(1-\alpha)\right) \\
          &= f(\alpha w + (1-\alpha)u) + \frac{\lambda\alpha}{2}||w||^2 + \frac{\lambda(1-\alpha)}{2}||u||^2 \\
          &\leq \alpha f(w) + (1-\alpha)f(u) + \frac{\lambda\alpha}{2}||w||^2 + \frac{\lambda(1-\alpha)}{2}||u||^2 \\
          &\leq \alpha (f(w) + \frac{\lambda}{2}||w||^2) + (1-\alpha)(f(u) + \frac{\lambda}{2}||u||^2) \\
          &= \alpha g(w) + (1-\alpha)g(u)
        \end{align*}
        as desired.

        Now we must prove the converse. To that aim, assume $g$ is $\lambda$-strongly convex. We will now prove that $f$ is convex.
        \begin{align*}
          &\hphantom{=} f(\alpha w + (1-\alpha)u) \\
          &= g(\alpha w + (1-\alpha)u) - \frac{\lambda}{2}||\alpha w + (1-\alpha)u||^2 \\
          &\leq \alpha g(w) + (1-\alpha)g(u) - \frac{\lambda}{2}||\alpha w + (1-\alpha) u||^2 - \frac{\lambda}{2}\alpha(1-\alpha)||w-u||^2 \\
          &\leq \alpha g(w) + (1-\alpha)g(u) - \frac{\lambda}{2}(\alpha w + (1-\alpha) u)^T(\alpha w + (1-\alpha) u) - \\
          &\hphantom{\leq} \frac{\lambda}{2}\alpha(1-\alpha)(w-u)^T(w-u) \\
          &\leq \alpha g(w) + (1-\alpha)g(u) - \frac{\lambda}{2}\left(\alpha^2 ||w||^2 + 2\alpha(1-\alpha)(u \cdot w) + (1-\alpha) ||u||^2\right) - \\
          &\hphantom{\leq} \frac{\lambda}{2}\alpha(1-\alpha)(||w||^2 - 2(u\cdot w) + ||u||^2) \\
          &\leq \alpha g(w) + (1-\alpha)g(u) - \frac{\lambda}{2}\left(\alpha^2 ||w||^2 + (1-\alpha) ||u||^2\right) - \frac{\lambda}{2}\alpha(1-\alpha)(||w||^2 + ||u||^2) \\
          &\leq \alpha g(w) - \frac{\lambda}{2}||w||^2\left(\alpha^2 + \alpha(1-\alpha) \right)  + (1-\alpha)g(u) - \frac{\lambda}{2}||u||^2\left((1-\alpha)^2 + \alpha(1-\alpha)\right) \\
          &\leq \alpha g(w) - \frac{\lambda\alpha}{2}||w||^2  + (1-\alpha)g(u) - \frac{\lambda(1-\alpha)}{2}||u||^2 \\
          &\leq \alpha \left(g(w) - \frac{\lambda}{2}||w||^2\right) + (1-\alpha)\left(g(u) - \frac{\lambda}{2}||u||^2\right) \\
          &= \alpha f(w) + (1-\alpha)f(u)
        \end{align*}
        so $f$ is convex.
  \end{enumerate}
\end{solution}

\begin{problem}{Exercise 3.6 (Random Variables and Strong Convexity)}
  Let $X \in \mathbb{R}$ be a discrete vector-valued random variable, with distribution
  \[
  \rho(X) =
  \begin{cases}
    w, & \text{with probability }\alpha \\
    u, & \text{with probability }(1 - \alpha)
  \end{cases}
  \]
  \begin{enumerate}
      \item Find $E[X]^2$, $E[X^2]$, $Var(X)$. Show that the expression $E[X^2] - (E[X])^2 = Var(X)$ corresponds to
      \[
      \alpha f(w) + (1 - \alpha) f(u) - f(\alpha w + (1 - \alpha) u) = \alpha (1 - \alpha) \|w - u\|^2
      \]
      with $f(w) = w^2$.

      \item Suppose $f$ is $\lambda$-strongly convex. Let $X$ be as above. Show that
      \[
      E[f(X)] - f(E[X]) \geq \frac{\lambda}{2} Var(X)
      \]
  \end{enumerate}
\end{problem}
